Physcraper: Body

### 2020-07-30 2-body.Rmd
ejm comments intro

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with new DNA sequence data, geographical location, and other metadata in a reproducible and continuous manner is possible by automating connections between biological databases. Here, we propose Physcraper, a tool that uses existing phylogenetic research and data from public biological databases to update a starting tree and single locus alignments, to build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.

@@

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multi-gene data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships [@song2012resolving].
Genomics has, and will continue to, revolutionize phylogenetic inference.
However, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing (RADSeq), single nucleotide polymorphisms (SNPs), microsatellites, and ultra-conserved elements (UCEs).
Having a variety of approaches can result in largely non-overlapping genomic data sets across taxa.
Also, as @andermann2020guide point out, "assembling full genomes is often unnecessary for phylogenomic studies if the main goal is to retrieve an appropriate number of phylogenetically informative characters from several independent and single-copy genetic markers [@jones2016targeted]."
While targeted capture approaches are ameliorating the problem of non-overlapping genomic data sets [@andermann2020guide; @jones2016targeted], in the meantime decades of single locus sequencing have already generated homologous DNA data that can be used for phylogenetic reconstruction at many scales.

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy at the time [about `r round(236023/240708, digits=4)*100`% of taxa; @sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand", but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
However, building molecular character matrices of few, thoughtfully curated markers, including manual curation of alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking these alignments with data from molecular databases that has not yet been included in any public phylogenetic estimate, has the potential to enrich and update phylogenetic relationships in many regions of the tree of life. The phylogenies associated to all these alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking these trees to their corresponding alignment repository in TreeBASE is available.
<!-- However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link. -->

We present Physcraper, a Python pipeline that relies on programmatic access protocols to link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to automatically update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for custom comparison of phylogenetic hypotheses.
Physcraper focuses on database interoperability, by integrating taxonomic name matching across biological databases. This integration makes downstream analyses straightforward, and allows users to place new data from a range biological database in an evolutionary context.

@@

- In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem]. If the user is not ready to make the input tree public, tree tip labels must be standardized to the unified OpenTree taxonomy using the bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool. *Should we saythat it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?*
  This step is referred to as taxon name mapping. If taxonomic names can't be mapped, their taxonomic information will be excluded from further analysis.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the taxonomic name mapping and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline. Currently, only trees connected to a published study can be stored in the Phylesystem.
  <!-- Discussion: (although there are plans to allow storage of unpublished trees). -->
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on
  TreeBASE.
  <!--
  **EJM: Currently 2950 studies have treebase data links. It is possible that some/many of these have multiple trees in them, but that is harder to count.** -->
  <!-- there are `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website
  ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)).   -->

>>>>

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with new DNA sequence data, geographical location, and other metadata in a reproducible and continuous manner is possible by automating connections between biological databases. Here, we introduce Physcraper, a tool that uses existing phylogenetic research from public biological databases to update a starting tree and single locus alignments, to build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.

@@

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multiple single locus data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships [@song2012resolving].
Genomics has, and will continue to, revolutionize phylogenetic inference.
Yet, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing, single nucleotide polymorphisms, microsatellites, and ultra-conserved elements, which has lead to largely non-overlapping genomic data sets across taxa, difficulting wide scale phylogenetic reconstructions.
<!-- Also, as @andermann2020guide point out, "assembling full genomes is often unnecessary for phylogenomic studies if the main goal is to retrieve an appropriate number of phylogenetically informative characters from several independent and single-copy genetic markers [@jones2016targeted]." -->
While phylogenomics ameliorates the problem of non-overlapping genomic data sets by focusing on targeted capture of informative characters from independent and single-copy genetic markers [@andermann2020guide; @jones2016targeted], decades of single locus sequencing have already generated homologous DNA data sets that can be used for phylogenetic reconstruction at many scales.

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy at the time [about `r round(236023/240708, digits=4)*100`% of taxa; @sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand", but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
However, building molecular character matrices of few, thoughtfully curated markers and alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking published alignments with data from molecular databases that has not yet been included in any public phylogenetic estimate, has the potential to enrich and update phylogenetic relationships in many regions of the tree of life. The phylogenies associated to TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.
<!-- However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link. -->

Physcraper relies on programmatic access protocols to automatically link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for custom comparison of phylogenetic hypotheses.
Physcraper is coded as a Python pipeline that focuses on database interoperability, by integrating taxonomic name matching across biological databases. This integration makes downstream analyses straightforward, and allows users to place new data from a range biological database in an evolutionary context.

@@

- In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem]. If the user is not ready to make the input tree public, tree tip labels must be standardized to the unified OpenTree taxonomy using the bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool. *Should we say that it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?*
  This step is referred to as taxon name mapping. If taxonomic names can't be mapped, their taxonomic information will be excluded from further analysis.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the taxonomic name mapping and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline. Currently, only trees connected to a published study can be stored in the Phylesystem.
  <!-- Discussion: (although there are plans to allow storage of unpublished trees). -->
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on
  TreeBASE.
  <!--
  **EJM: Currently 2950 studies have treebase data links. It is possible that some/many of these have multiple trees in them, but that is harder to count.** -->
  <!-- there are `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website
  ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)).   -->
